A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-11 sections/algorithm.tex

This section introduces a novel algorithmic system that implements the
DK's declarative specification. The algorithmic system comes with
several novelties.\bruno{fill in/summarize novelties. I think
  novelties include: precise scoping control; CPS-style; dealing with
  inference judgements.}

@@

\paragraph{Existential Variables} In order to find the unknown type $\tau$ in the declarative system,
\emph{existential variables} $\al, \bt$ are introduced in the algorithmic system.
They behave similarly to unification variables,
but their scoping is restricted by the position located in the
worklist\bruno{be careful to ensure that the reader knows about
  worklist at this point},
rather than globally defined.
Any existential variable $\al$ should only be solved to
a well-formed type under the worklist before $\al$ is defined.
Thus the mono-type $\tau$ represented by the corresponding existential variable
is always well-formed under the corresponding declarative context.
In other words, the position of $\al$'s reflects the well-formedness restriction of $\tau$'s.

@@

\paragraph{Judgments and judgment chains} 
A \emph{judgment chain} $\jg$ is a core component of algorithmic
type-checking. For conciseness, in the rest of the paper, we usually
refer to ``judgment chains'' simply by ``judgments''.  There are four kinds
of judgments in our system: subtyping ($A \le B$), checking ($e\Lto
A$), inference ($e\To_{a} \j$) and
application inference ($\appInfAlg{A}{e}$).  Subtyping and checking are relatively simple,
since the result could be either success or fail, and no other output
is returned.  However inference and application inference return some
type that leads to further computation. We use a
continuation-passing-style encoding for them.  For example, a judgment
chain $e \To_a (a \le B)$ contains two judgments: first we want to
infer the type of the expression $e$, and then check if that type is a
subtype of $B$. The \emph{unknown} type of $e$ is represented by a
type variable $a$, which is used in the second judgement to denote the 
type of $e$.

\paragraph{Worklist Judgments} Our algorithm has a non-standard form.
We combine the traditional contexts and judgment(s) into a single worklist.
The worklist collects variable bindings and judgments in \emph{order}, which represents their scopings:
each variable is visible only after it is declared.
For example, $[a,x:a,x \Lto a]$ is a valid worklist, but $[\textcolor{red}{x} \Lto a, x:\textcolor{red}{a},a]$ is not (red-colored symbols are out-of-scope).

@@

The algorithmic typing reduction rules, defined in Figure~\ref{fig:alg}, have the form $\Gm \rto \Gm'$.
The worklists $\Gm, \Gm'$ contain both variable declarations and judgments.
In every step reduction pops the first judgment from the worklist for processing,
possibly pushes new judgments to the worklist, and repeats this process until the list is empty.
A worklist that finally reduces to empty is considered to be successfully type-checked.
When the first element is a variable declaration, we simply continue
the process (Rules 1-3).
Note that those rules about to garbage collecting variables that are
out-of-scope for the next judgement being processed.
The other rules always deal with the first judgment in the worklist. Logically we can discern 6 group of rules.

Firstly, we have six rules (Rules 4-9) that are similar to those of
declarative subtyping.  For instance, Rule 7 consumes one subtyping
judgment and pushes two back to the worklist.  Rule 8 differs from
declarative Rule ${\le}{\forall}L$ by introducing an existential
variable $\al$ instead of guessing the monotype $\tau$ for
instantiation.  Each existential variable is later solved to a
monotype $\tau$ with the same context, so the final solution of
$\al$, $\tau$, should be well-formed under context $\Gm$.

\paragraph{Variable scopings within worklist}
Rules 8 and 9 involve variable declaration, which demonstrates how our
worklist treats variable scopings. Rule 8 introduces an existential
variable $\al$ that is only visible to the judgment $[\al/a]A \le B$.
Then reduction continues until all the subtyping judgments after $\al$
are satisfied.  Finally we can safely remove $\al$ since no occurrence
of $\al$ could leak to its left in the worklist.  On the other hand,
removing $\al$ according to the algorithm garbage-collects the
variable at the perfect time: it leaves the environment immediately
after being unreferenced.\bruno{Perhaps this is better illustrated
  with a simple example?}

@@

\paragraph{Existential decomposition}
Secondly, Rules 10 and 11 are algorithmic extensions to Rule $\mathtt{{\le}{\to}}$.
Both of them partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We remove $\al$ from the worklist, and insert $\al[1], \al[2]$ back to the same position.
To propagate the instantiation to the rest of the worklist and maintain the well-formedness,
every $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition is to prevent a diverging infinite loop.
Without such check, $\al \le 1 \to \al$, as an example, would diverge.

\paragraph{Solving existentials.} Thirdly, Rules 12-17 are base cases where an existential variable is solved.
Each of the rules removes and existential variable, substitutes the
variable with its solution and continues. Importantly the rules
respect the scoping of existential variables. For example, rule 12 
states that an existential variable $\al$ can be solved with another
existential variable $\bt$ only if $\bt$ occurs after $\al$.

\paragraph{Checking judgements}
Fourthly, Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, and rewrited in the
continuation-passing-style.
\bruno{The side condition needs to be explained, and some mention of
  the overlapping is probably needed here.}
Rules 19 and 20 adapts the worklist style from their corresponding declarative rules.
Rule 21 is a special case of $\mathtt{Decl\to I}$.\bruno{Rule 21
  deserves some more explanation. What is the motivation for this
  rule? You may want to give some concrete example to motivate it.}
Instantiation technique similar to Rules 10 and 11 applies.
The declarative checking rule $\mathtt{Decl1I}$ is not translated to the algorithm,
since combining Rule 18 and Rule 24 would give the same result.

\paragraph{Inference judgments}
Fifthly, Rules 22-26 deal with inference judgments.
Inference judgment chains are written in continuation-passing-style.
When an inference judgment succeeds with type $A$,
the algorithm continues to work on the inner-chain $\jg$ by assigning $A$ to its placeholder varible $a$.
Rule 23 infers an annotated expression by changing into checking mode,
therefore another judgment chain is created.
Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$.
The variable scopings are well-maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
\bruno{You have some note ``differs'' on the rules. Should that
  deserve an explanation here?}

@@

\subsection{Multiple reduction}
The syntax $\Gm\redto \Gm'$ denotes multiple reduction.
If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully.

>>>>

This section introduces a novel algorithmic system that implements 
DK's declarative specification. The algorithmic system comes with
several novelties.\bruno{fill in/summarize novelties. I think
  novelties include: precise scoping control; CPS-style; dealing with
  inference judgements.}

@@

\paragraph{Existential Variables} In order to find unknown types $\tau$ in the declarative system,
the algorithmic system introduces \emph{existential variables} $\al, \bt$.
They behave similarly to unification variables,
but their scoping is restricted by their position in the
worklist\bruno{be careful to ensure that the reader knows about
  worklist at this point},
rather than being global.
Any existential variable $\al$ should only be solved to
a type that is well-formed with respect to the worklist to which $\al$ has been added.
The point is that the mono-type $\tau$ represented by the corresponding existential variable
is always well-formed under the corresponding declarative context.
In other words, the position of $\al$'s reflects the well-formedness restriction of $\tau$'s.

@@

\paragraph{Judgments and judgment chains} 
\emph{Judgment chain}s $\jg$ are the core components of our algorithmic
type-checking. For conciseness, in the rest of the paper, we usually
refer to ``judgment chains'' simply by ``judgments''.  There are four kinds
of judgments in our system: subtyping ($A \le B$), checking ($e\Lto
A$), inference ($e\To_{a} \jg$) and
application inference ($\appInfAlg{A}{e}$).  Subtyping and checking are relatively simple,
since their result is only success or failure. However both inference and
application inference return a type that is used in further computation. We use
continuation-passing-style encoding to accomplish this. For example, the judgment
chain $e \To_a (a \le B)$ contains two judgments: first we want to
infer the type of the expression $e$, and then check if that type is a
subtype of $B$. The \emph{unknown} type of $e$ is represented by a
type variable $a$, which is used in the second judgement to denote the 
type of $e$.

\paragraph{Worklist Judgments} Our algorithm has a non-standard form.
We combine the traditional contexts and judgment(s) into a single worklist.
The worklist collects variable bindings and judgments in \emph{order}, which represents their scopings:
each variable is visible only after it is declared.
For example, $[a,x:a,x \Lto a]$ is a valid worklist, but $[\textcolor{red}{x} \Lto a, x:\textcolor{red}{a},a]$ is not (the red-colored symbols are out of scope).

@@

The algorithmic typing reduction rules, defined in Figure~\ref{fig:alg}, have the form $\Gm \rto \Gm'$.
The worklists $\Gm, \Gm'$ contain both variable declarations and judgments.
The reduction process treats the worklist as a stack.
In every step it pops the first judgment from the worklist for processing,
possibly pushes new judgments onto the worklist, and repeats this process until the list is empty.
A worklist that fully reduces to empty corresponds to a successful type-checking.

Rules 1-3 pop a variable declaration; this garbage collects variables that are
out of scope for the remaining judgements.

All other rules concern a judgement at the front of the worklist. Logically we can discern 6 group of rules.

Firstly, we have six subtyping rules (Rules 4-9) that have similar 
declarative counterparts.  For instance, Rule 7 consumes a subtyping
judgment and pushes two back to the worklist.  Rule 8 differs from
declarative Rule ${\le}{\forall}L$ by introducing an existential
variable $\al$ instead of guessing the monotype $\tau$ for
instantiation.  Each existential variable is later solved to a
monotype $\tau$ with the same context, so the final solution $\tau$ of
$\al$ should be well-formed under context $\Gm$.

\paragraph{Variable scopings within worklist}
Rules 8 and 9 involve variable declarations and demonstrate how our
worklist treats variable scopings. Rule 8 introduces an existential
variable $\al$ that is only visible to the judgment $[\al/a]A \le B$.
Reduction continues until all the subtyping judgments after $\al$
are satisfied.  Finally we can safely remove $\al$ since no occurrence
of $\al$ could have leaked into the left part of the worklist.  Moreover,
the algorithm garbage-collects the $\al$
variable at the perfect time: it leaves the environment immediately
after being unreferenced.\bruno{Perhaps this is better illustrated
  with a simple example?}\tom{Yes, it would be good to show one or more
example derivations to illustrate the key points.}

@@

\paragraph{Existential decomposition}
Secondly, Rules 10 and 11 are algorithmic extensions to Rule $\mathtt{{\le}{\to}}$.
Both of them partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We remove $\al$ from the worklist, and insert $\al[1], \al[2]$ back to the same position.
To propagate the instantiation to the rest of the worklist and maintain the well-formedness,
every $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition is to prevent a diverging infinite loop.
For example, without such check $\al \le 1 \to \al$ would diverge.

\paragraph{Solving existentials.} Thirdly, Rules 12-17 are base cases where an existential variable is solved.
Each of the rules removes an existential variable and substitutes the
variable with its solution in the remaining worklist. Importantly the rules
respect the scoping of existential variables. For example, rule 12 
states that an existential variable $\al$ can be solved with another
existential variable $\bt$ only if $\bt$ occurs after $\al$.

\paragraph{Checking judgements}
Fourthly, Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, but rewritten in the
continuation-passing-style.
\bruno{The side condition needs to be explained, and some mention of
  the overlapping is probably needed here.}
Rules 19 and 20 adapt their declarative counterparts to the worklist style.
Rule 21 is a special case of $\mathtt{Decl\to I}$.\bruno{Rule 21
  deserves some more explanation. What is the motivation for this
  rule? You may want to give some concrete example to motivate it.}
The same instantiation technique as in Rules 10 and 11 applies.
The declarative checking rule $\mathtt{Decl1I}$ is not translated to the algorithm,
since combining Rule 18 and Rule 24 would give the same result.

\paragraph{Inference judgments}
Fifthly, Rules 22-26 deal with inference judgments.
Inference judgment chains are written in continuation-passing-style.
When an inference judgment succeeds with type $A$,
the algorithm continues to work on the inner-chain $\jg$ by assigning $A$ to its placeholder variable $a$.
Rule 23 infers an annotated expression by changing into checking mode,
therefore another judgment chain is created.
Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$.
The variable scopings are maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
\bruno{You have some note ``differs'' on the rules. Should that
  deserve an explanation here?}

@@

\subsection{Multiple reduction}
The syntax $\Gm\redto \Gm'$ denotes multiple reduction steps.
If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully.
